Viewing Computer Science through Citation Analysis

### 2019-12-20 Scientometrics/comp.tex
minor edits, but look at the last paragraph

\begin{abstract}
Computer science has experienced dramatic growth and diversification over the last twenty years. Towards a current understanding of the structure of this discipline, we analyze 
a cohort of the computer science literature using the DBLP database. For insight on the features of this cohort and the relationship within its components, we constructed article 
level clusters based on either direct citations or co-citations, and reconcile them to major and minor subject categories in the All Science Journal Classification (ASJC). 
We described complementary insights from clustering by direct citation and co-citation, and both point to the increase in computer science publications and their scope.
Our analysis shows cross-category clusters, some that interact with external fields, such as the biological sciences, while others remain inward looking.

@@

 \emph{Overview} For the purpose of this study, our working definition of the computer science literature is all publications in the DBLP bibliography that (i) had a digital object identifier (DOI), and (ii) could also be matched to article identifiers in the Scopus bibliography. The underlying assumption is that records in DBLP are greatly enriched for computer science even while recognizing that coverage may be incomplete and possibly biased. Cross-matching DBLP publications to records in the Scopus abstract and citation database of peer-reviewed literature enabled us to harvest the richer links in Scopus as well as extract links to publications from other disciplines. This cross-matching to Scopus also allows the use of both journal and article based classifications when clustering documents. We used DBLP articles from journals and conference proceedings to construct article clusters by using either direct citations or co-citations as links. We reconciled these clusters with the All Science Journal Classification (ASJC) developed and maintained by Scopus through a combination of automated and manual procedures, producing a dataset of  2,685,356 publications.

@@

 \emph {Direct Citation} Graclus~\cite{graclus_2007} is a spectral graph clustering package that optimizes  various clustering criteria, including normalized cut, ratio cut, and ratio association, and that has previously been applied to citation data~\cite{subelj_clustering_2016}. 
We used v1.2 in our experiments. The \emph{comp} dataset was formatted as an undirected graph, stored in a file with a header line indicating the number of nodes and edges, and used as input to Graclus, which requires the number of clusters to be formed as an input parameter. In preliminary experiments, we varied the number of clusters to be formed between 10 and 50 clusters (data not shown).  At around 20 clusters, clusters size was relatively stable with the largest cluster containing roughly 10 times the number of nodes in the smallest one~\cite{subelj_clustering_2016}. Consequently, we used Graclus to generate 20 clusters, labeled 0-19 (Table~\ref{tab:graclus}), analogous to Level 1 of Waltman and van Eck's mapping of nearly 10 million publications but focused on the DBLP bibliography rather than a broader Web of Science sample~\cite{waltman_new_2012}.

@@

In our implementation of variable level clustering (Fig.~\ref{fig:quad-chart}(a)), we first calculated the number of citations accumulated across Scopus for all 2,685,356 articles in \emph{comp}. After discarding those publications without any citations, we restricted further analysis to highly cited articles--those in the 90th percentile of citations or higher--resulting in a dataset of size 212,311. We then identified 4,318,305 publications in Scopus that cite these 212,311 articles from \emph{comp}. For each of the 4,318,305 citing publications in turn, all possible ${n \choose 2}$ reference pairs were generated from a publication's cited references, where $n$ is the number of references in a publication. The cited reference pairs this generated were then restricted to those in the set of 212,311 highly cited papers previously identified. A total of 46,463,117 unique co-cited pairs were thus obtained. 
The frequency of these co-cited pairs was then computed across the \emph{comp} dataset  and normalized using Salton's cosine formula~\cite{salton_citation_1979} to limit dominance by areas with high citation activity. These data were represented in a graph where each node was a publication and the weighted edge between the pair was the normalized co-citation frequency. \par

We applied variable clustering with initial parameters as follows: the threshold  $t$  is initially set to  the median normalized co-citation frequency (quantile=0.5), increment $i = 0.1$, and maximum cluster size, $mcs=200$. Thus, at the start, all edges below the median normalized co-citation frequency were deleted. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5--0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster.  Using this approach, 22,232 clusters containing 84,591 nodes were generated, each containing less than 100 nodes. Clusters containing only two nodes were discarded, bringing the total number of clusters down to 10,298. The publications in these 10,298 clusters were overwhelmingly drawn from the Physical Sciences (one of the four top level categories in the Scopus ASJC classification), of which computer science is a sub-category (Fig.~\ref{fig:quad-chart}).

@@

These data suggest that areas central to computer science in 2019 (Salton and Bergmark's historical triad of hardware, software, and theory) are more likely to be found in multiple clusters than peripheral topics. A second inference is that, in some cases, journal-based classification and our article clusters align fairly well (Hardware and Architecture). 
A third inference is that the ASJC minor subject area ``Computer Applications" is relatively broad, and publications thus labeled are present at the $>=$ 15\% level in 16 out of 20 clusters.  Finally,   for this DBLP dataset, as we clustered it, interactions with fields outside computer science such as Biology are (Biochemistry, Neurology)  are  detected in two separate clusters. The first appears to be the interaction of Biochemistry, Molecular Biology, and Genetics with Statistics, Mathematics, and Computer Science, and the second is the interaction of Neurology, Cognitive Neuroscience, and Artificial Intelligence.

@@

Clustering by co-citation begins, for each cluster, with a pair of nodes that nucleates its subsequent formation. Thus, we also designated the pair of nodes in a cluster with the strongest edge as its nucleating pair and labeled the cluster by inspecting this pair of documents and assigning an intuitive label to it. These labels show high degree of correspondence  to the ASJC minor subject area that accounts for the largest fraction of the nodes in a co-citation cluster (Table~\ref{tab:centroid_reconcile}). 
The extent of this correspondence %between the manually applied label and the ASJC minor subject area 
suggests that using the nucleating cluster as the basis for labeling co-citation clusters (as we generated them) may be valid, although its usefulness is likely to vary according to the data being examined. We provide the DOIs of these nucleating pairs matched to manually assigned labels for independent review (Table~\ref{tab:centroid}).

@@

\section{Discussion} Considering expansion and diversification of the field of computer science, we revisited its characterization in 1979~\cite{salton_citation_1979}. In comparison to Salton and Bergmark, we analyzed considerably more data, 2.68 million publications versus 391 by using two bibliographic databases, DBLP and Scopus, consisting of  approximately 7 million and 88 million publications respectively. By linking the two datasets to harvest citation and metadata, we were able to construct article level clusters in two different ways and reconcile these clusters with the Scopus ASJC classification, which provides a journal level perspective.

@@

A comparison of the clusters generated using direct citation and co-citation shows interesting contrasts. The nodes in a co-citation cluster often map largely to one or two direct citation clusters, for example. 98\% of the nodes in co-citation cluster 19008 map to direct citation cluster 8, which in turn aligns with theory, software, applications, and networks. Conversely, co- citation cluster 18947 nucleated by a pair of articles in the Journal of Applied Mathematics and Computation is distributed between direct clusters 11,16, and 17 effectively spanning Artificial Intelligence, Applied Mathematics, Biochemistry, Chemistry, Genetics, Theory, Software, Applications, and Statistics.

The focus of this article was on high-level features, and a preference for simplicity and intuitiveness in the choice of methods. This preference may address questions  regarding generalizability and the specific choices we made, such as the (i) use of the DBLP dataset matched to Scopus, which may not capture all aspects of computer science and its interactions with other fields, (ii) the basis for clustering, and (iii) mapping the results of this clustering against a classification designed around journals rather than individual articles. We believe that a journal level combined with article level approach is useful for studies of this kind.

We speculate that biology, often dominant in bibliometric studies, is restricted to two clusters on account of (i) the focus of DBLP (ii) our use of normalized co-citations (ii) the threshold set for detection. Future research should, of course, include complementary investigations at finer levels of granularity and sensitivity using article-level approaches that others have developed.

In the 40 years since Salton and Bergmark's landmark paper, the field of computer science has not only expanded in volume, it has expanded in its interactions with other fields, and has
also resulted in new disciplinary and interdisciplinary subfields.

>>>>

\begin{abstract}
Computer science has experienced dramatic growth and diversification over the last twenty years. Towards a current understanding of the structure of this discipline, we analyze 
a cohort of the computer science literature using the DBLP database. For insight on the features of this cohort and the relationship within its components, we constructed article 
level clusters based on either direct citations or co-citations, and reconciled them to major and minor subject categories in the All Science Journal Classification (ASJC). 
We described complementary insights from clustering by direct citation and co-citation, and both point to the increase in computer science publications and their scope.
Our analysis shows cross-category clusters, some that interact with external fields, such as the biological sciences, while others remain inward looking.

@@

 \emph{Overview} For the purpose of this study, our working definition of the computer science literature was all publications in the DBLP bibliography that (i) had a digital object identifier (DOI), and (ii) could also be matched to article identifiers in the Scopus bibliography. The underlying assumption is that records in DBLP are greatly enriched for computer science even while recognizing that coverage may be incomplete and possibly biased. Cross-matching DBLP publications to records in the Scopus abstract and citation database of peer-reviewed literature enabled us to harvest the richer links in Scopus as well as extract links to publications from other disciplines. This cross-matching to Scopus also allowed the use of both journal- and article-based classifications when clustering documents. We used DBLP articles from journals and conference proceedings to construct article clusters by using either direct citations or co-citations as links. We reconciled these clusters with the All Science Journal Classification (ASJC) developed and maintained by Scopus through a combination of automated and manual procedures, producing a dataset of  2,685,356 publications.

@@

 \emph {Direct Citation} Graclus~\cite{graclus_2007} is a spectral graph clustering package that optimizes  various clustering criteria, including normalized cut, ratio cut, and ratio association, and that has previously been applied to citation data~\cite{subelj_clustering_2016}. 
We used v1.2 in our experiments. The \emph{comp} dataset was formatted as an undirected graph, stored in a file with a header line indicating the number of nodes and edges, and used as input to Graclus, which requires the number of clusters to be formed as an input parameter. In preliminary experiments, we varied the number of clusters to be formed between 10 and 50 clusters (data not shown).  At around 20 clusters, clusters size was relatively stable with the largest cluster containing roughly 10 times the number of nodes in the smallest one,
so that 20 clusters is a good choice with respect to the criteria specified in~\cite{subelj_clustering_2016}. Consequently, we used Graclus to generate 20 clusters, labeled 0-19 (Table~\ref{tab:graclus}), analogous to Level 1 of Waltman and van Eck's mapping of nearly 10 million publications but focused on the DBLP bibliography rather than a broader Web of Science sample~\cite{waltman_new_2012}.

@@

We first calculated the number of citations accumulated across Scopus for all 2,685,356 articles in \emph{comp}. After discarding those publications without any citations, we restricted further analysis to highly cited articles--those in the 90th percentile of citations or higher--resulting in a dataset of size 212,311. We then identified 4,318,305 publications in Scopus that cite these 212,311 articles from \emph{comp}. For each of the 4,318,305 citing publications in turn, all possible ${n \choose 2}$ reference pairs were generated from a publication's cited references, where $n$ is the number of references in a publication. The cited reference pairs this generated were then restricted to those in the set of 212,311 highly cited papers previously identified. A total of 46,463,117 unique co-cited pairs were thus obtained. 
The frequency of these co-cited pairs was then computed across the \emph{comp} dataset  and normalized using Salton's cosine formula~\cite{salton_citation_1979} to limit dominance by areas with high citation activity. These data were represented in a graph where each node was a publication and the weighted edge between the pair was the normalized co-citation frequency. \par

In our implementation of variable level clustering (Fig.~\ref{fig:quad-chart}(a)), 
we set initial parameter values as follows: the threshold  $t$  is initially set to  the median normalized co-citation frequency (quantile=0.5), increment $i = 0.1$, and maximum cluster size, $mcs=200$. Thus, at the start, all edges below the median normalized co-citation frequency were deleted. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5--0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster.  Using this approach, 22,232 clusters containing 84,591 nodes were generated, each containing less than 100 nodes. Clusters containing only two nodes were discarded, bringing the total number of clusters down to 10,298. The publications in these 10,298 clusters were overwhelmingly drawn from the Physical Sciences (one of the four top level categories in the Scopus ASJC classification), of which computer science is a sub-category (Fig.~\ref{fig:quad-chart}).

@@

These data suggest that fields central to computer science in 2019 (Salton and Bergmark's historical triad of hardware, software, and theory) are more likely to be found in multiple clusters than peripheral fields. A second inference is that, in some cases, journal-based classification and our article clusters align fairly well (Hardware and Architecture). 
A third inference is that the ASJC minor subject area ``Computer Applications" is relatively broad, and publications thus labeled are present at the $>=$ 15\% level in 16 out of 20 clusters.  Finally,   for this DBLP dataset, as we clustered it, interactions with fields outside computer science such as Biology  (i.e., Biochemistry, Neurology)  are  detected in two separate clusters. The first appears to be the interaction of Biochemistry, Molecular Biology, and Genetics with Statistics, Mathematics, and Computer Science, and the second is the interaction of Neurology, Cognitive Neuroscience, and Artificial Intelligence.

@@

Clustering by co-citation begins, for each cluster, with a pair of nodes that nucleates its subsequent formation. Thus, we also designated the pair of nodes in a cluster with the strongest edge as its nucleating pair and labeled the cluster by manually labeling the nucleating pair of documents.  
These labels show a high degree of correspondence  to the ASJC minor subject area that accounts for the largest fraction of the nodes in a co-citation cluster (Table~\ref{tab:centroid_reconcile}). 
%The extent of this correspondence %between the manually applied label and the ASJC minor subject area 
Thus, using the nucleating pair as the basis for labeling co-citation clusters (as we generated them) may be valid, although its usefulness is likely to vary according to the data being examined; we also note that manual annotation (although beneficial) is not scalable. We provide the DOIs of these nucleating pairs matched to manually assigned labels for independent review (Table~\ref{tab:centroid}).

@@

\section{Discussion} 
Considering expansion and diversification of the field of computer science, we revisited its characterization by Salton and Bergmark in 1979~\cite{salton_citation_1979}. In comparison to \cite{salton_citation_1979}, we analyzed considerably more data, 2.68 million publications versus 391 by using two bibliographic databases, DBLP and Scopus, consisting of  approximately 7 million and 88 million publications respectively. By linking the two datasets to harvest citation and metadata, we were able to construct article level clusters in two different ways and reconcile these clusters with the Scopus ASJC classification, which provides a journal level perspective.

@@

A comparison of the clusters generated using direct citation and co-citation shows interesting contrasts. 
The nodes in a co-citation cluster often map largely to one or two direct citation clusters: for example, 98\% of the nodes in co-citation cluster 19008 map to direct citation cluster 8, which in turn aligns with theory, software, applications, and networks. 
Conversely, co-citation cluster 18947 nucleated by a pair of articles in the Journal of Applied Mathematics and Computation is distributed between direct citation clusters 11, 16, and 17, effectively spanning Artificial Intelligence, Applied Mathematics, Biochemistry, Chemistry, Genetics, Theory, Software, Applications, and Statistics.

The focus of this article was on high-level features, and a preference for simplicity and intuitiveness in the choice of methods. This preference may address questions  regarding generalizability and the specific choices we made, such as the (i) use of the DBLP dataset matched to Scopus, which may not capture all aspects of computer science and its interactions with other fields, (ii) the basis for clustering, and (iii) mapping the results of this clustering against a classification designed around journals rather than individual articles. We believe that an approach that combines journal-level  with article-level analyses is useful for studies of this kind.

We speculate that biology, often dominant in bibliometric studies, is restricted to two clusters on account of (i) the focus of DBLP,  (ii) our use of normalized co-citations, and (ii) the threshold set for detection. Future research should, of course, include complementary investigations at finer levels of granularity and sensitivity using article-level approaches that others have developed.

In the 40 years since Salton and Bergmark's landmark paper, the field of computer science has not only expanded in volume, it has expanded in its interactions with other fields, and has
also resulted in new disciplinary and interdisciplinary subfields. 
Furthermore, machine learning and data science, which build off computer science and statistics, 
are emerging as major fields that are driving innovation in industry and science, and research in these areas is increasingly being performed in
fields external to computer science.
While DBLP provides an insight into what is commonly accepted as computer science, additional evaluation of the broader literature that
uses and develops computer science is needed to better assess the impact of computer science.
